---
title: "Climate events impact on population health and economy, 1950-2011"
output: 
  html_document:
    code_folding: "show"
    keep_md: true
    css: css_for_assignment.css
    toc: true
    number_sections: TRUE
---

# Summary

<style>
div.blue { background-color : #7FB3D5; color : #ffffff; border-radius: 15px; padding: 15px 10px 10px 10px;}
</style>
<div class = "blue">

This report describes the impact of climate events on population health and the economy through the years from 1950 to 2011. The main goal of it is to identify which events are the most harmful. After exploratory analysis, I found out that the most dangerous for human health events are hurricane/typhoon, extreme heat, and tsunami and for the economy - hurricane/typhoon, storm surge, and severe thunderstorm.

</div>

# Data Processing

First of all, I prepared an environment for myself - uploaded all needed libraries (please check if you have all of them installed before proceeding) and created a theme for *ggplot* graphs.


```{r message=FALSE}

library(tidyverse)
library(extrafont)
library(scales)
library(lubridate)
library(xtable)
library(knitr)
library(kableExtra)
library(DT)
library(psych)

violet = '#D98880'
blue = '#31698a'
lightblue = '#7FB3D5'
pink = '#F2D7D5'

theme <- theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, 
                                    size = 16, 
                                    family = "Kodchasan", 
                                    colour = violet,
                                    face = "bold"),
          plot.subtitle = element_text(hjust = 0.5),
          text = element_text(family = "Montserrat"),
          legend.position = "bottom",
          strip.background = element_rect(fill = "white"),
          strip.text = element_text(colour = blue))

```

Secondly, I uploaded the data directly from the course website, saved it in *data* directory under my working directory and uploaded it to R as *raw_data*. If such directory didn't exist, the code would create it for me. 

```{r message=FALSE, cache=TRUE}

data_folder <- "data"

if (!file.exists(data_folder)){
    dir.create(data_folder)
}

download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              "data/data.csv.bz2")

raw_data <- read_csv("data/data.csv.bz2")

```

Let's check what the data looks like and if we have any unusual values.

```{r}

glimpse(raw_data)
summary(raw_data)

```

After seeing the structure of the data,  I decided to change the type of `BNG_DATE`, and `BNG_TYPE` as both of them were imported as character vectors and have them as `date` and `time` (plus `date_time` as their combination). Also, I already noticed that the columns for population health (`FATALITIES` and `INJURIES`) are full so no need to worry about empty values. The last thing I did was converted column names with *tolower()* and removed the only line with without proper event type value.

```{r}

proc_data <- raw_data %>%
    mutate(date = date(mdy_hms(BGN_DATE)),
           time = hms::as.hms(paste(substr(BGN_TIME, 1, 2), substr(BGN_TIME, 3, 2), '00', sep = ":")),
           date_time = date + time,
           propexp = case_when(PROPDMGEXP == "K" ~ 1000,
                               PROPDMGEXP == "M" ~ 1000000,
                               PROPDMGEXP == "B" ~ 1000000000,
                               TRUE ~ 1),
           cropexp = case_when(CROPDMGEXP == "K" ~ 1000,
                               CROPDMGEXP == "M" ~ 1000000,
                               CROPDMGEXP == "B" ~ 1000000000,
                               TRUE ~ 1)) %>%
    select(-BGN_TIME, -BGN_DATE) %>%
    filter(EVTYPE != "?")

names(proc_data) <- tolower(names(proc_data))

head(proc_data)

```

# Overall stats

First I check how many events were measured every year. From the graph below, I can see that the amount of events is increasing exponentially and the main growth started in 1994. 

```{r fig.align="center"}

proc_data %>%
    group_by(year = year(date)) %>%
    summarize(num = n()) %>%
    ggplot(aes(x = year,
               y = num)) +
    geom_col(fill = blue) +
    scale_y_continuous(labels = comma_format()) +
    scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
    labs(title = "Yearly amount of events",
         x = "Date",
         y = "Events") +
    theme

```

To understand what kind of plotting possibilities I have, I have checked how many different event types there are:

```{r}

unique_types <- length(unique(proc_data$evtype))

```

`r unique_types` seems like a really big amount of factors to use in the plot so I decided to go with tables instead.

# Results

## Population Health

I checked the impact on population health by averaging amount of injuries and fatalities on each event. In the table below, I provided only the events with at least 10 occurrences in total and arranged by the total impact on population health.

```{r}

data_for_table <- proc_data %>%
    group_by(`Event Type` = evtype) %>%
    summarize(`Total Events` = n(),
              `Average Fatalities` = round(mean(fatalities, na.rm = TRUE), 2),
              `Average Injuries` = round(mean(injuries, na.rm = TRUE), 2),
              `Average Total Health` = round(mean(injuries + fatalities, na.rm = TRUE), 2)) %>%
    arrange(desc(`Average Total Health`)) %>%
    filter(`Total Events` >= 10)

datatable(data_for_table, rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T))
    
```

The most dangerous for human health events are hurricane/typhoon, extreme heat, and tsunami.

## Economy

I checked the impact on the economy by averaging the damage by each event. In the table below, I provided only the events with at least 10 occurrences in total and arranged by the total impact on the economy.

```{r}

data_for_table <- proc_data %>%
    group_by(`Event Type` = evtype) %>%
    summarize(`Total Events` = n(),
              `Mean Flood Damage (K USD)` = round(mean(propexp * propdmg, na.rm = TRUE) / 1000, 2),
              `Mean Crop Damage (K USD)` = round(mean(cropexp * cropdmg, na.rm = TRUE) / 1000, 2),
              `Mean Damage (K USD)` = round(mean(propexp * propdmg + cropexp * cropdmg, na.rm = TRUE) / 1000, 2)) %>%
    arrange(desc(`Mean Damage (K USD)`)) %>%
    filter(`Total Events` >= 10)

datatable(data_for_table, rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T))

```

The biggest damage on the economy is done by hurricane/typhoon, storm surge, and severe thunderstorm.
